---
title: "R Packages"
format: html
---

# R Packages for Data Science

In R, packages extend core functionality. Below we cover installation, loading, and examples for essential packages.

## 1. Installing and Loading Packages

```{r}
# Install packages from CRAN
install.packages("tidyverse")   # collection for data manipulation and visualization
install.packages("vegan")       # ecological and multivariate analysis
install.packages("readr")       # fast data import
install.packages("sf")          # spatial data

# Load packages
library(tidyverse)
library(vegan)
library(readr)
library(sf)
```

## 2. tidyverse

The **tidyverse** is a suite by Hadley Wickham that shares an underlying design philosophy: - **dplyr**: Grammar of data manipulation (`filter`, `select`, `mutate`, `summarise`, `arrange`). - **ggplot2**: Grammar of graphics for layered plotting. - **tidyr**: Tools to reshape and tidy data. - **readr**: Fast and friendly data import.

### Example: Data Wrangling and Plotting

```{r}
# Load iris as tibble
tibble_iris <- as_tibble(iris)

# Compute mean petal length by species
tibble_iris %>%
  group_by(Species) %>%
  summarise(mean_petal = mean(Petal.Length), sd_petal = sd(Petal.Length))

# Plot distribution of Sepal.Length
ggplot(tibble_iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  labs(title = "Sepal Length by Species", y = "Sepal Length (cm)")
```

## 3. vegan

**vegan** is specialised for community ecology: - **diversity()**: Calculate diversity indices (Shannon, Simpson). - **rda() / cca()**: Redundancy analysis, canonical correspondence. - **metaMDS()**: Non-metric multidimensional scaling.

### Example: Ecological Analysis

```{r}
# Load dune dataset from vegan
data(dune)
data(dune.env)

# Compute Shannon diversity for each site
diversities <- diversity(dune)
diversities

# Run NMDS ordination
nmds <- metaMDS(dune)
plot(nmds)
```

## 4. Additional Packages

-   **sf**: Simple features for spatial data. Use `st_read()` to load shapefiles.
-   **lubridate**: Work with dates and times (`ymd`, `hms`).
-   **stringr**: Consistent functions for string manipulation.
-   **purrr**: Functional programming tools (map, reduce).

Explore CRAN or use `browseURL('https://cran.r-project.org/web/packages/')` to find more.
