---
title: "R Basics"
format: html
---

# R Basics

Welcome to a foundational lesson on base R functions, datasets, and data exploration. Treat this as your first lecture: we will cover how to compute descriptive statistics, access built-in datasets, and view data within RStudio.

## 1. Descriptive Statistics in Base R

Without loading any packages, R provides functions to compute common statistics directly:

```{r}
# Create a numeric vector
values <- c(12.5, 17.8, 9.3, 14.1, 11.7)

# Arithmetic mean (average)
mean_value <- mean(values)
mean_value

# Standard deviation (population default)
sd_value <- sd(values)
sd_value

# Summary of vector (min, 1Q, median, mean, 3Q, max)
summary(values)
```

**Explanation**: - `mean(x)`: sums all elements of `x` and divides by the count. - `sd(x)`: calculates the square root of the variance (with denominator `n-1`). - `summary(x)`: returns a six-number summary plus the mean.

You can also compute other measures: - **Median**: `median(values)` - **Variance**: `var(values)` - **Quantiles**: `quantile(values, probs = c(0.25, 0.5, 0.75))` - **Range**: `range(values)`

## 2. Built-in Datasets in R

R includes numerous datasets for practice. To see all available datasets, run:

```{r}
# List all datasets
data()
```

Below are some key datasets you will encounter:

-   **iris**: Fisher's iris data (1936). Measures of sepal & petal lengths for three species of iris. 150 observations, 5 variables.
-   **mtcars**: Motor Trend Car Road Tests (1974). Fuel consumption and 10 aspects of automobile design and performance for 32 automobiles.
-   **airquality**: New York Air Quality Measurements (1973). Daily readings of ozone, solar radiation, wind, and temperature in New York.
-   **USArrests**: Violent Crime Rates by US State (1973). Murder, Assault, UrbanPop, and Rape statistics for 50 states.
-   **PlantGrowth**: Results of an experiment on plant weight under different treatments.

To load and inspect a dataset:

```{r}
# Load iris and view first rows
data(iris)
head(iris)

# Summary of mtcars
data(mtcars)
summary(mtcars)
```

## 3. Viewing Data in RStudio

You have two main options:

1.  **View() function**: Opens a spreadsheet-like viewer.

```{r}
View(iris)
```

2.  **Environment pane**: In RStudio, click the dataset name under **Environment** → **Data**. You can sort, filter, and search within this view.

Use these tools to familiarise yourself with your data before analysis.
