---
title: "R Basics"
format: html
---

# R Basics

Welcome to your first R lecture. We will explore fundamental functions for descriptive statistics, discover R’s built-in datasets, and practise viewing data in RStudio.

## 1. Computing Descriptive Statistics

Base R provides the following core functions without loading any extra packages:

```{r}
# Sample numeric vector
data_vec <- c(8.2, 5.4, 9.7, 6.1, 7.3, 8.9)

# Compute descriptive statistics
mean_val   <- mean(data_vec)      # arithmetic mean
sd_val     <- sd(data_vec)        # standard deviation
median_val <- median(data_vec)    # median
var_val    <- var(data_vec)       # variance
quants     <- quantile(data_vec, probs = c(0.25, 0.5, 0.75))  # quartiles
range_val  <- range(data_vec)     # min and max

# Display results
mean_val
sd_val
median_val
var_val
quants
range_val
```

### Explanation of Functions

-   `mean(x)`: average of the elements in `x`.
-   `sd(x)`: sample standard deviation (denominator *n*−1).
-   `median(x)`: middle value of sorted `x`.
-   `var(x)`: sample variance.
-   `quantile(x, probs)`: specified percentiles of `x`.
-   `range(x)`: minimum and maximum values of `x`.

Use `?function_name` (e.g., `?mean`) in the console to view documentation.

## 2. Built-in Datasets in R

R comes with many datasets. List them all with:

```{r}
# Show available datasets
data()
```

Below are commonly used examples:

```{r}
# Inspect iris dataset
data(iris)
head(iris)
summary(iris)
str(iris)
```

| Dataset | Description | Dimensions |
|--------------|--------------------------------------------|--------------|
| `iris` | Edgar Anderson’s measurements of iris flowers | 150 × 5 |
| `mtcars` | Motor Trend car specifications and fuel consumption | 32 × 11 |
| `airquality` | Daily air quality measurements in New York (1973) | 153 × 6 |
| `CO2` | Carbon dioxide uptake in grass plants under different conditions | 84 × 5 |
| `PlantGrowth` | Results of an experiment on plant weight under control and treat. | 30 × 2 |

## 3. Viewing Data in RStudio

-   **View( )**: In the console or script, run:

    ```{r}
    View(iris)
    ```

    This opens a spreadsheet-like viewer.

-   **Environment Pane**: Under the **Environment** tab, click a dataset name to inspect it interactively. Use the gear icon ⚙️ to show/hide columns.

Always check dimensions (`dim(x)`), variable names (`names(x)`), and types (`class(x)` / `typeof(x)`) before proceeding.
